Generating Correlated Data for Omics Simulation

### 2025-08-07 manuscript/paper.qmd
revisions for editorial and reviewer comments

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    margin-left: 0.75in
    margin-right: 0.75in
    margin-top: 0.75in
    margin-bottom: 0.75in
bibliography: references.bib
---
## Abstract

@@

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
See @fig-intro a for a high-level overview.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing the covariance matrix $\Sigma_{\text{sim}}$ to use during simulation presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{\text{ref}}$ of the reference dataset (possibly after some transformation of the data).
However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covariance matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.

@@

We present three methods - to be called PCA, spiked Wishart, and corpcor - for making the choice of $\Sigma_{\text{sim}}$
All three take the form
$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
An additional advantage of this special form is that data may be generated very efficiently (@fig-intro b,c) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.

@@

For comparison we also ran SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see Supplemental Methods for details.

Simulated data captures the genes' mean and variance accurately (@fig-compare-to-real a-b).
Next, we compared to the real dataset when projected onto the top two principal components of the real dataset (@fig-compare-to-real c).
The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.

Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods modestly improved upon the independent simulation.
The SPsimSeq simulator performed similarly to the PCA method.

Lastly, we compared the variances of principal components on each dataset (@fig-compare-to-real e).
These were computed separately for each dataset, unlike (@fig-compare-to-real c) which used the reference dataset's PCA weights for all datasets.
The independent data has much lower variance than the real dataset in the top four principal components.
The spiked Wishart method comes closest to the real dataset, as it optimizes for fitting these values.
Surprisingly, the corpcor method performs only somewhat better than the independent method.
The PCA method puts a large amount of variance into the first two components (due to using $k=2$) and then undershoots the other components.
Like PCA, the SPsimSeq simulator, overshot the first PCA component but did a better job matching lower PCA coordinates.

@@

We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 1.

@@

We used a fly whole body RNA-seq dataset GSE81142, see Supplemental Figure 2.
To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
This was repeated 20 times for each of four dependence methods (independent, PCA, Wishart, and corpcor).

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
While the mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.

To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e,f) as for the fly whole body datasets.

@@

We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 3, which contains a total of 77 samples every 3 hours, for 36 hours.
As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples.
We computed the means of gene expressions at each time point.
We then used these to simulate 20 time series datasets for each of the four simulation methods: independent, PCA, Wishart, and corpcor.

@@

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS accuracy depended significantly on this parameter, with the default producing good accuracy across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
At that setting, CYCLOPS accuracy is much higher in the PCA method and moderately improved in Wishart method, compared to the 'independent' method. 
This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
Notably, the number of eigengenes and accuracy tracked both with each other and with the amount of gene-gene correlation (see Supplemental Figure 3 d) across the methods.
This demonstrates that the correlation structure of the transcriptome can have a major impact on accuracy.

@@

Other Gaussian copula-based R packages include `copula`, `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases, likely due to the reliance on the Cholesky decomposition, see @fig-intro d.

@@

### Supporting Information 1

Supporting document containing all Supplemental Figures 1-3 and Supplemental Methods.

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
crossref:
    fig-prefix: Fig
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    margin-left: 0.75in
    margin-right: 0.75in
    margin-top: 0.75in
    margin-bottom: 0.75in
bibliography: references.bib
---
## Abstract

@@

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
See [@fig-intro]A for a high-level overview.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing the covariance matrix $\Sigma_{\text{sim}}$ to use during simulation presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{\text{ref}}$ of the reference dataset (possibly after some transformation of the data).
However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covariance matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see [@fig-intro]B.

@@

We present three methods - to be called PCA, spiked Wishart, and corpcor - for making the choice of $\Sigma_{\text{sim}}$
All three take the form
$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
An additional advantage of this special form is that data may be generated very efficiently ([Figs @fig-intro]B and [-@fig-intro]C) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.

@@

For comparison we also ran SPsimSeq, another RNA-seq simulator for both bulk and single-cell that uses a Gaussian copula-based model of dependence, see S1 Text for details.

Simulated data captures the genes' mean and variance accurately ([Figs @fig-compare-to-real]A and [-@fig-compare-to-real]B).
Next, we compared to the real dataset when projected onto the top two principal components of the real dataset ([@fig-compare-to-real]C).
The simulations with dependence are distributed around the entire space like the real data, but the independent simulations have unrealistically low variance in these components, clustering tightly around the origin.

Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
The simulation with independence showed the least levels of gene-gene correlations ([@fig-compare-to-real]D).
However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods modestly improved upon the independent simulation.
The SPsimSeq simulator performed similarly to the PCA method.

Lastly, we compared the variances of principal components on each dataset ([@fig-compare-to-real]E).
These were computed separately for each dataset, unlike ([@fig-compare-to-real]C) which used the reference dataset's PCA weights for all datasets.
The independent data has much lower variance than the real dataset in the top four principal components.
The spiked Wishart method comes closest to the real dataset, as it optimizes for fitting these values.
Surprisingly, the corpcor method performs only somewhat better than the independent method.
The PCA method puts a large amount of variance into the first two components (due to using $k=2$) and then undershoots the other components.
Like PCA, the SPsimSeq simulator, overshot the first PCA component but did a better job matching lower PCA coordinates.

@@

We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see S1 Fig.

@@

We used a fly whole body RNA-seq dataset GSE81142, see S2 Fig.
To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
This was repeated 20 times for each of four dependence methods (independent, PCA, Wishart, and corpcor).

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed ([Figs @fig-DESeq2]A-[-@fig-DESeq2]D).
We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
While the mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.

To demonstrate the application of our simulation method for another organism, we also simulated datasets using the mouse cortex dataset GSE151923 [@Wang2022] and selected samples from male mice, as in @fig-compare-to-real.
We then simulated differential expression experiments as above and observed a similar result ([Figs @fig-DESeq2]E and [-@fig-DESeq2]F) as for the fly whole body datasets.

@@

We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
We chose a mouse cortex time series dataset GSE151565, see S3 Fig, which contains a total of 77 samples every 3 hours, for 36 hours.
As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples.
We computed the means of gene expressions at each time point.
We then used these to simulate 20 time series datasets for each of the four simulation methods: independent, PCA, Wishart, and corpcor.

@@

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS accuracy depended significantly on this parameter, with the default producing good accuracy across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its accuracy depends significantly on the dependence structure of the simulated time series data ([@fig-CYCLOPS]A).
At that setting, CYCLOPS accuracy is much higher in the PCA method and moderately improved in Wishart method, compared to the 'independent' method. 
This difference is likely driven by the difference in the number of eigengenes used ([@fig-CYCLOPS]B), which is a measure of how much dependence is present in the dataset.
Notably, the number of eigengenes and accuracy tracked both with each other and with the amount of gene-gene correlation (see S3D Fig) across the methods.
This demonstrates that the correlation structure of the transcriptome can have a major impact on accuracy.

@@

Other Gaussian copula-based R packages include `copula`, `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods which scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
All of these packages provide more flexibility in specifying dependence and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.
Another package that optimizes the speed of random multivariate normal vector generation is the `mvnfast` package for R [@mvnfast], however, we show that this does not provide a meaningful speedup for our use-cases, likely due to the reliance on the Cholesky decomposition, see [@fig-intro]C.

@@

### S1 Fig

Evaluation of our simulation on a metabolomics dataset.

### S2 Fig

Evaluation of our simulation on a fly whole-body RNA-seq dataset.

### S3 Fig

Evaluation of our simulation on a mouse cortex RNA-seq dataset.

### S1 Text

Supplemental methods.

### 2025-08-07 manuscript/supplemental/supplemental.qmd
Revise according to editorial requests
revisions for editorial and reviewer comments

```{r}
#| label: fig-compare-to-real-fly
#| fig-cap: "Comparison to real data run on a fly whole body data set from GSE81142. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 8
#| fig-height: 8
readRDS("../../processed/compare_to_real_plot.GSE81142.RDS")
```

```{r}
#| label: fig-compare-to-real-mouse
#| fig-cap: "Comparison to real data run on a mouse cortex timeseries body data set from GSE151565. Only timepoint ZT0 was used for this comparison. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 8
#| fig-height: 8
readRDS("../../processed/compare_to_real_plot.GSE151565.RDS")
```
\FloatBarrier
## Supplemental Methods

@@

>>>>

```{r}
#| label: fig-compare-to-real-fly
#| fig-cap: "Comparison to real data run on a fly whole body data set from GSE81142. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 8
#| fig-height: 8
readRDS("../../processed/compare_to_real_plot.GSE81142.RDS")
```

```{r}
#| label: fig-compare-to-real-mouse
#| fig-cap: "Comparison to real data run on a mouse cortex timeseries body data set from GSE151565. Only timepoint ZT0 was used for this comparison. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 30 reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data."
#| fig-width: 8
#| fig-height: 8
readRDS("../../processed/compare_to_real_plot.GSE151565.RDS")
```
\FloatBarrier
## Supplemental Methods

@@

### Running mvnfast

The `mvnfast` R package provides a faster implementation of the multivariate normal distribution random vector generator function, called `rmvn`. 
As for `mvrnorm`, we set means to be the mean expression of each gene and used the sample covariance matrix as the `Sigma` parameter.
One core was used for a fair comparison to the other methods, though `mvnfast` does support multithreaded execution.
Note that `mvnfast` has an option for using a provided pre-computed Cholesky decompositoin of the covariance matrix instead of producing it itself.
Since Cholesky decomposition is the primary time-sink, using this option could likely considerably speed up the execution time if such a decomposition is known.
We did not use this option since the comparison is for a generic covariance matrix, where the Cholesky decomposition is not known a priori.
This contrasts to our implementation which avoids Cholesky decomposition by using only covariance matrices of a specific form.
